Normalization by Evaluation for Sized Dependent Types

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] First run over section 3 done.

@@

Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory and
figures~\ref{fig:ty} and \ref{fig:eq} the inference rules.

@@

%Our type theory is formulated via the following judgements, which are mutually inductively defined.

%In addition, we will refer to the following derived judgements.

% Inference rules.

% \fbox{$\der \Gamma$} (implies $\der \resurrect \Gamma$).
% \[
%   \ru{}{\der \cempty}
% \qquad
%   \ru{\der \Gamma \qquad
%       \resurrect \Gamma \der T
%     }{\der \cext \Gamma T}
% \qquad
%   \ru{\der \Gamma
%     }{\der \cext \Gamma \Size}
% \qquad
%   \ru{\der \Gamma
%     }{\der \erext \Gamma \Size}
% \]

% \fbox{$\Gamma(i) = \eann T$} (implies $\resurrect\Gamma \der T$ if\/ $\der \Gamma$).
% \[
%   \ru{}{(\eext \Gamma T)(0) = \eann T \slift}
% \qquad
%   \ru{\Gamma(i) = \eann T
%     }{(\cext \Gamma \_)(i+1) = \eann T \slift}
% \]

%Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

% \fbox{$\Gamma \der t = t' : T$} (implies $\resurrect \Gamma \der T$ and $\Gamma \der t,t' : T$ [and $\resurrect\Gamma \der t = t' : T$].)

% Liftings: $\fbox{$\xi : \Gamma \leq \Delta$}
%   \defas \Gamma \der \xi : \Delta \mand \xi = \slift^k_m \mforsome k,m$
% %  \defas \xi = \slift^k_m \mwith \Gamma \der \xi : \Delta$
% (implies $\der \Gamma$ and $\der \Delta$).

% (implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \xi : \Delta$).
% \begin{gather*}
%   \ru{\der \Gamma
%     }{\sempty : \Gamma \leq \cempty}
% \qquad
%   \ru{\xi : \Gamma \leq \Delta \qquad
%       \resurrect\Gamma \der T
%     }{\xi \slift : \cext \Gamma T \leq \Delta}
% \qquad
%   \ru{\xi : \Gamma \leq \Delta \qquad
%       \resurrect \Delta \der U \qquad
%       \resurrect\Gamma \der T \leq U \xi
%     }{(\xi \slift, \ind 0) : \cext \Gamma T \leq \cext \Delta U}
% \end{gather*}

% Substitution typing. % and equality.
% \fbox{$\Gamma \der \sigma : \Delta$}
% (implies $\der \Gamma$ and $\der \Delta$
%  and $\resurrect\Gamma \der \sigma : \Delta$
%  and $\resurrect\Gamma \der \sigma : \resurrect \Delta$).
% % and
% % \fbox{$\Gamma \der \sigma = \sigma' : \Delta$}.

% \begin{gather*}
%   \ru{\der \Gamma \qquad
%     }{\Gamma \der \sempty : \cempty}
% \qquad
%   \ru{\Gamma \der \sigma : \Delta \qquad
%       \resurrect \Delta \der T \qquad
%       \Gamma \der t : T \sigma
%     }{\Gamma \der \sext \sigma t : \cext \Delta T}
% \qquad
%   \ru{\Gamma \der \sigma : \Delta \qquad
%       \Gamma \der a : \Size
%     }{\Gamma \der \sext \sigma a : \eext \Delta \Size}
% \end{gather*}

% \begin{gather*}
%   \ru{\der \Gamma \qquad
%     }{\Gamma \der \sempty = \sempty : \cempty}
% \qquad
%   \ru{\Gamma \der \sigma = \sigma': \Delta \qquad
%       \resurrect \Delta \der T \qquad
%       \Gamma \der t = t' : T \sigma
%     }{\Gamma \der \sext \sigma t = \sext {\sigma'} t': \cext \Delta T}
% \end{gather*}

% Substitution typing and equality.
% \fbox{$\suty \Gamma \sigma \tau \Delta$} and
% \begin{gather*}
%   \ru{\der \Gamma \qquad
%     }{\suty \Gamma \sempty \sempty \cempty}
% \qquad
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \resurrect \Delta \der T \qquad
%       \Gamma \der u = t : T \tau
%     }{\suty \Gamma {\sext \sigma u} {\sext \tau t} \cext \Delta T}
% \\[2ex]
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \Gamma \der a = b : \Size
%     }{\suty \Gamma {\sext \sigma a} {\sext \tau b} \cext \Delta \Size}
% \qquad
%   \ru{\suty \Gamma \sigma \tau \Delta \qquad
%       \Gamma \der a, b : \Size
%     }{\suty \Gamma {\sext \sigma a} {\sext \tau b} \erext \Delta \Size}
% \end{gather*}

@@

In the following, let $J$ %, $L$, and $R$
match a part of a judgement.

@@

\begin{lemma}[Resurrection]
\label{lem:res} \bla
\begin{enumerate}
\item $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
  % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
  % Further, if\/ $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
\item If\/ $\Gamma \der J : \resurrect \Delta$ % and $\der \Delta$
  then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der \sigma : \Delta$ % then $\Gamma \der \sigma : \resurrect\Delta$.
  then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
\end{enumerate}
\end{lemma}

@@

\begin{lemma}[Substitution equality]
\label{lem:subeq}\bla
\begin{enumerate}
\item Conversion: If\/ $\sueq \Gamma \sigma {\sigma'} {\tau_1} \Delta$ and $\sueq {\resurrect\Gamma} {\tau_1} {\tau_2} {\tau} {\resurrect\Delta}$ then $\sueq \Gamma \sigma {\sigma'} {\tau_2} \Delta$.
\item Reflexivity: If\/ $\Gamma \der \sigma : \Delta$ then $\sueq \Gamma \sigma \sigma \sigma \Delta$.
\item Symmetry:  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then $\sueq \Gamma {\sigma'} {\sigma} {\tau} \Delta$.
\item Transitivity:  If\/ $\sueq \Gamma {\sigma_1} {\sigma_2} {\tau} \Delta$ and $\sueq \Gamma {\sigma_2} {\sigma_3} {\tau} \Delta$ then $\sueq \Gamma {\sigma_1} {\sigma_3} {\tau} \Delta$.
\item Functionality: Let $\sueq \Gamma \sigma {\sigma'} \tau \Delta$.
  \begin{enumerate}
  \item
  If\/ $\Delta \der t : T$ then $\Gamma \der t\sigma = t\sigma' : T\tau$.
  \item
  If\/ $\Delta \der t = t' : T$ then $\Gamma \der t\sigma = t'\sigma' : T\tau$.
  \item
  Corollary: If\/ $\Delta \der T \leq T'$ then $\Gamma \der T\sigma \leq T\sigma'$.
  \end{enumerate}
\end{enumerate}
\end{lemma}
\begin{corollary}[Partial resurrection for substitution equality]
  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then
  $\sueq {\resurrect \Gamma} \tau \tau \tau {\resurrect \Delta}$.
\end{corollary}

>>>>

% For having two versions of a paper in one file.
% Stuff that does not fit into the short version can be encosed in \LONGVERSION{...}
\ifdefined\LONGVERSION
  \relax
\else
% short version:
\newcommand{\LONGVERSION}[1]{}
\newcommand{\SHORTVERSION}[1]{#1}
% % long version:
% \newcommand{\LONGVERSION}[1]{#1}
% \newcommand{\SHORTVERSION}[1]{}
% \newcommand{\SHORTVERSION}[1]{BEGIN~SHORT\ #1 \ END~SHORT}
\fi
\newcommand{\LONGSHORT}[2]{\LONGVERSION{#1}\SHORTVERSION{#2}}
\newcommand{\SHORTLONG}[2]{\SHORTVERSION{#1}\LONGVERSION{#2}}
\newcommand{\EXTENDED}[1]{}

@@

@@

Fig.~\ref{fig:judge} lists the inductive and derived judgements of our type theory and
figures~\ref{fig:ty} and \ref{fig:eq} the inference rules.
In the typing judgement $\Gamma \der t : T$, the term $t$ is in scope of
$\Gamma$, \ie, may not mention irrelevant variables in relevant
positions.  However, the type $T$ is in scope of the resurrected
context $\resurrect\Gamma$, hence, can mention all variables declared
in $\Gamma$.  The other judgements are organized similarly.  To
understand this distinction, consider judgement
$i \div \Size \der \Nat i$.  This would mean that $i$ is irrelevant in
$\Nat i$ and thus, $\Gamma \der \Nat a = \Nat {a'}$ for all sizes
$\resurrect\Gamma \der a,a' : \Size$.  But this is exactly wrong!  However,
judgement $i \div \Size \der \zero i : \Nat (i + 1)$ is fine, it
implies $\Gamma \der \zero a = \zero {a'} : \Nat (b+1)$ for all
$\resurrect\Gamma \der a,a',b : \Size$.

Our substitution theorem
needs to reflect the distinct scope of things left of the colon vs.\
things right of the colon.  In the last example we have applied the
substitution triple $\Gamma \der [a] = [a'] \eeq [b] : (i \div \Size)$
to judgement $i \div \Size \der \zero i : \Nat (i + 1)$.
The first two substitutions apply to the term side while the third
substitution applies to the type side.  The fact that we replace an
irrelevant variable $i$ allows $a,a',b$ to refer to irrelevant
variables from $\Gamma$, thus, they are in scope of $\resurrect\Gamma$.

@@

In the following, we list syntactic properties of our judgements.  To
this end, let $J$ %, $L$, and $R$
match a part of a judgement.

@@

All types in a context are considered in the resurrected context,
which justifies the first statement of the following lemma.
A resurrected context is more permissive, as it brings more variable
into scope.  As such, it is comparable to an extended context or a
context where types have been replaced by subtypes.  This intuition
accounts for the remaining statments but (\ref{it:ressubst}).  The
latter is a defining property of substitutions: only replacement for
irrelevant sizes may refer to irrelevant size variables.
\begin{lemma}[Resurrection]
\label{lem:res} \bla
\begin{enumerate}
\item \label{it:reswf} $\der \Gamma$ iff $\der \resurrect\Gamma$;  then $\resurrect \Gamma \der \sid : \Gamma$.
  % Consequently, if\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
  % Further, if\/ $\Gamma \der J : \resurrect\Delta$ then $\Gamma \der J : \Delta$.
\item If\/ $\Gamma \der J$ then $\resurrect \Gamma \der J$.
\item If\/ $\Gamma \der J : \resurrect \Delta$ % and $\der \Delta$
  then $\Gamma \der J : \Delta$.
\item \label{it:ressubst} If\/ $\Gamma \der \sigma : \Delta$ % then $\Gamma \der \sigma : \resurrect\Delta$.
  then $\resurrect\Gamma \der \sigma : \resurrect\Delta$.
\end{enumerate}
\end{lemma}

@@

The relation $\sueq \Gamma \sigma {\sigma'} \tau \Delta$ is a partial equivalence
relation (PER) on term-side substitutions $\sigma,\sigma'$. Note that
we cannot resurrect this judgement to
$\sueq{\resurrect\Gamma}{\sigma}{\sigma'}\tau \Delta$ in general!
The type-side substitution $\tau$ can be converted to one that is
equal in and for resurrected contexts.
\begin{lemma}[Substitution equality]
\label{lem:subeq}\bla
\begin{enumerate}
\item Conversion: If\/ $\sueq \Gamma \sigma {\sigma'} {\tau_1} \Delta$ and $\sueq {\resurrect\Gamma} {\tau_1} {\tau_2} {\tau} {\resurrect\Delta}$ then $\sueq \Gamma \sigma {\sigma'} {\tau_2} \Delta$.
\item Reflexivity: If\/ $\Gamma \der \sigma : \Delta$ then $\sueq \Gamma \sigma \sigma \sigma \Delta$.
\item Symmetry:  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then $\sueq \Gamma {\sigma'} {\sigma} {\tau} \Delta$.
\item Transitivity:  If\/ $\sueq \Gamma {\sigma_1} {\sigma_2} {\tau} \Delta$ and $\sueq \Gamma {\sigma_2} {\sigma_3} {\tau} \Delta$ then $\sueq \Gamma {\sigma_1} {\sigma_3} {\tau} \Delta$.
\item Functionality: Let $\sueq \Gamma \sigma {\sigma'} \tau \Delta$.
  \begin{enumerate}
  \item
  If\/ $\Delta \der t : T$ then $\Gamma \der t\sigma = t\sigma' : T\tau$.
  \item
  If\/ $\Delta \der t = t' : T$ then $\Gamma \der t\sigma = t'\sigma' : T\tau$.
  \item
  Corollary: If\/ $\Delta \der T \leq T'$ then $\Gamma \der T\sigma \leq T\sigma'$.
  \end{enumerate}
\end{enumerate}
\end{lemma}

\LONGVERSION{
\begin{corollary}[Partial resurrection for substitution equality]
  If\/ $\sueq \Gamma \sigma {\sigma'} {\tau} \Delta$ then
  $\sueq {\resurrect \Gamma} \tau \tau \tau {\resurrect \Delta}$.
\end{corollary}
}
